---
title: "Assignment 9"
output: html_notebook
---

Using the dataset from last week, test a model that postulates that threat
mediates the relationship between empathy and prejudice (see the scripts from
the last two weeks to determine which indicators define those variables).

Install packages:
```{r}
library(pacman)
p_load(semPlot, tidyverse, lavaan, psych, kableExtra)
```
Load data:
```{r}
data <- read.delim("../assignment_9/data/Northdale survey _reverse coded_.dat")
```
Convert the variables:
```{r}
#   Threat
    data$Thrt_Phy1<-data$Thrt_Phy1/3
    data$Thrt_Phy2<-data$Thrt_Phy2/3
    data$Thrt_Phy3<-data$Thrt_Phy3/3
#   Prejudice
    data$Att_IS1<-(data$Att_IS1-1)/6.75    
    data$Att_IS2<-(data$Att_IS2-1)/6.75  
    data$Att_IS4<-(data$Att_IS4-1)/6.75  
#   Contact quantity
    data$Cnt_pos_B<-data$Cnt_pos_B/2
    data$Cnt_pos_IS1<-data$Cnt_pos_IS1/2
#   Contact_quality
    data$Cnt_Qul_IS1<-data$Cnt_Qul_IS1/3
    data$Cnt_Qul_IS2<-data$Cnt_Qul_IS2/3
    data$Cnt_Qul_IS3<-data$Cnt_Qul_IS3/3
#   Empathy
    data$Empath1<-data$Empath1/3    
    data$Empath2<-data$Empath2/3
    data$Empath3<-data$Empath3/3
```
```{r}
path.dat <- data
```
Test the model:
```{r}
#   SEM of model1
    pathm1 <- '
            # MEASUREMENT MODEL
              threat_scale       =~ Thrt_Phy1 + Thrt_Phy2 + Thrt_Phy3
              prejudice_scale    =~ Att_IS1 + Att_IS2 + Att_IS4
              empathy_scale      =~ Empath1 + Empath2 + Empath3
      
            # STRUCTURAL MODEL
              threat_scale ~ empathy_scale
              prejudice_scale ~ threat_scale + empathy_scale
'
```
```{r}
prejpathfit1 <- sem(pathm1, data = path.dat)
    summary(prejpathfit1 , fit.measures = T, standardized = T)
```
The model that suggests that empathy predicts prejucdice, both directly and through the mediator variable threat, is not significantly different from the saturated model (p = 0.26, df = 24). This means that it is not significantly different from the model that utilises all possible paths, despite it being simpler - a good indicator of parsimony. The model (model 1) is significantly different from the null model (p < 0.001, df = 36), indicating that we have not dropped too many potential paths, and our predictions are still better than the no-information rate. CFI and TLI - measures of our model's location on the continuum from saturated to null model - indicate that our model is far closer to the saturated model. THese measures should approximate 1, and values > 0.95 indicate a good fit. By these measures our model has a good fit (CFI = 1.00, TLI = 1.00). AIC and BIC are noted for comparison with other models. RMSEA value should approximate 0, and values < 0.05 indicate a good fit. By this measure our model has a good fit, with the upper end of the CI only slightly above optimal (RMSEA = 0.02, CI = 0.00 - 0.06). SRMR should approximate 0, and values < 0.05 indicate a good fit. This measure also indicates that our model has good fit (SRMR = 0.03).

An analysis of our latent variables indicates that all manifest variables significantly load onto the proposed latent variables.

Our regression analysis indicates that empathy does not significantly predict threat (beta = 0.05, p = 0.42). However, threat and empathy are both significant direct predictors of prejudice (empathy: beta = -0.24, p < 0.001; threat: beta = 0.34, p < 0.001). Increased empathy leads to decreased prejudice, and increased threat leads to increased prejuice (this second relationship is stronger).

All our etimates are positive, and the standard errors are relatively small.
-------------------------
Tabulate the direct, indirect and total effects of empathy on prejudice, and provide
significance tests where appropriate. See whether you can compute the p values
using normal theory and bootstrapping.
```{r}
#   SEM of model1 (direct and indirect effects)
    pathm1_effects <- '
            # MEASUREMENT MODEL
              threat_scale       =~ Thrt_Phy1 + Thrt_Phy2 + Thrt_Phy3
              prejudice_scale    =~ Att_IS1 + Att_IS2 + Att_IS4
              empathy_scale      =~ Empath1 + Empath2 + Empath3
  

          # direct effect
             prejudice_scale ~ c*empathy_scale
           
          # mediator
             threat_scale ~ a*empathy_scale
             prejudice_scale ~ b*threat_scale
           
          # indirect effect (a*b)
             ab := a*b
           
          # total effect
             total := c + (a*b)
         '
prejpathfit1_effects <- sem(pathm1_effects, data = path.dat)
summary(prejpathfit1_effects , fit.measures = T, standardized = T)
```
```{r}
colnames <- c("Value", "Direct effect", "Indirect effect", "Total effect")
betavalues <- c("std.all", -0.24, 0.02, -0.23)
pvalues <- c("p (>|z|)", 0.001, 0.425, 0.001)
effects_tab <- rbind(colnames, betavalues, pvalues)

```
```{r}
effects_tab %>%
  kable() %>%
  kable_styling()
```

